\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-10-24 demystify.tex
updates

    So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
    While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines.
    Furthermore, although many examples are specifics for George Mason University (GMU), the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} institutions in the US  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

@@

    I wish you the best of luck. And if you follow this guidance, you will at least have a good chance at GMU (see
    \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU}{why
    you want to study at GMU}). Happy school hunting!

@@

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low rankings from these schools (and be overlooked by Admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement and ask your LoR writers to talk about it.

@@

\paragraph{GRE} Most good CS PhD programs in the US \textbf{no longer require GREs}, so you don't need to take them. However, they might be useful for international students from programs we are not familiar with.

\begin{commentbox}[Vu:]
Personally I do not pay attention to GRE scores as I do not believe they are good indicator for research. However, if you have good GRE scores then you should include them in your application.
\end{commentbox}

@@

Choosing a school and an advisor is clearly one of the most important things in your mind when you apply to and especially when you get admitted multiple programs.  This is further complicated due to cultural differences and unfamiliarity of  international students to the US system and faculty.  This section aims to mitigate some confusions and help you make more informed decision.

@@

Most universities the US will have CS programs. 
However, while many universities offer Ph.D. degree in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD).

Thus, if your goal is Ph.D. in CS, you will have to filter out schools that do not offer such a degree.  This might sound obvious but many students (even those in the US) get confused due to the large number of universities. For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiplication locations of the University of Texas, e.g., Austin, Dallas, Arlington, all have PhDs in CS.

While this can be confusing at first, a little research, e.g., searching for PhD in CS from the school website, will help you find out. All schools listed in \S\ref{sec:ranking} have PhD studies in CS, so you can start there.

@@

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top schools that they dream about, and (ii) everything else.  In many cases, they use resources such as rankings from US News, which are \emph{highly} questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students evaluate CS programs using the reputations non-CS programs such as medical, math, or physics.
They even rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

@@

\begin{commentbox}[Dat] Most Vietnamese students, including those from top schools, \textbf{do not know} about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

@@

\begin{commentbox}[Xiaokuan]
Chinese students often only look at USNews rankings when selecting their Ph.D. universities (I did that, too, when I was applying for Ph.D. positions).
Now that I am a professor, I find it to be the least promising way.
%
The reason is that USNews does not provide a good metric for evaluating the quality of the Ph.D. program.
%
If you want to do great research, CSRankings is the best way to find good and active professors (which did not exist when I was applying),
since it solely focuses on publications at top-tier CS conferences.
%
Also, 
I think Ph.D. is not only about research; 
you need to also consider your daily life there, since you will (probably) stay for at least five years.
%
You might regret it if you did not consider this seriously before applying.
%
\end{commentbox}

@@

@@

\paragraph{Referring to professors you know} Here, because you are already familiar with these individuals, you can just informally use their firstname as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I did my postdoc with Jeff Foster at Univ. of Maryland.   Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many international students include a lengthy title of people they know, e.g., my advisor is Asst. Prof. X, and I also collaborate with Distinguished Scientist Y.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. In short, these nuances represent some cultural differences that you may encounter and will gradually adapt to. More on cultural differences in \S\ref{sec:cultural}.

>>>>

    So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
    While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines and should be 
    Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines.} institutions in the US  (though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU).

@@

    I wish you the best of luck. And if you follow this guidance, you will at least have a good chance at GMU (see
    \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU}{why
    you want to study at here}). Happy school hunting!

@@

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low rankings from these schools (and be overlooked by the admission committee).
So what to do with this? well, same as \hyperref[sec:your-school]{before}, e.g., put a note about this in your statement and ask your LoR writers to talk about it.

@@

\paragraph{GRE} Most good CS PhD programs in the US \textbf{no longer require GREs}, so you don't need to take them. However, they might be useful for international students from programs we are not familiar with.  If you have good GRE scores then you should include them in your application.

@@

Choosing a school and an advisor is clearly one of the most important things in your mind when you apply to and especially when you get admitted.  This is further complicated due to cultural differences and unfamiliarity of  international students to the education system in the .  This section aims to mitigate some confusions and help you make informed decision.

@@

Most universities the US will have CS programs. 
However, while many universities offer Ph.D. in their CS programs, many do not.  These universities might offer just Bachelor degrees (e.g., BS) and no graduate studies (i.e., no MS or Ph.D degrees), or they just offer MS programs (but no PhD). For example, Penn State in University Park has Ph.D. in CS,  but Penn State in Harrisburg only has BS and MS in CS, and Penn State in York only has BS in CS.  On the other hand, multiplication locations of the University of Texas, e.g., Austin, Dallas, Arlington, all have PhDs in CS.

Thus, if your goal is Ph.D. in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities. 
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. All schools listed in \S\ref{sec:ranking} have PhD studies in CS, so you can start there.

@@

International students not familiar with US universities often put them into \emph{two} bins:  (i) very top schools that they dream about, and (ii) everything else.  In many cases, they use resources such as rankings from US News, which are not very transparent and \emph{highly} questionable\footnote{\url{https://cra.org/cra-statement-us-news-world-report-rankings-computer-science-universities/}}.  Sometimes these students evaluate CS programs using the reputations non-CS programs such as medical, math, or physics.
They even rank universities based on popular states they know in the US, e.g., California and New York.  Clearly, there are so many thing wrong with these methods.

@@

\begin{commentbox}[Dat:] Most Vietnamese students, including those from top schools, \textbf{do not know} about CSRankings.  May be applicants who worked at top research places such as VinAI would know about it.
\end{commentbox}

@@

\begin{commentbox}[Xiaokuan:]
Chinese students often only look at USNews rankings when selecting their Ph.D. universities (I did that, too, when I was applying for Ph.D. positions).
Now that I am a professor, I find it to be the least promising way.
%
The reason is that USNews does not provide a good metric for evaluating the quality of the Ph.D. program.
%
If you want to do great research, CSRankings is the best way to find good and active professors (which did not exist when I was applying),
since it solely focuses on publications at top-tier CS conferences.
%
Also, 
I think Ph.D. is not only about research; 
you need to also consider your daily life there, since you will (probably) stay for at least five years.
%
You might regret it if you did not consider this seriously before applying.
%
\end{commentbox}

@@

\begin{center}
    \includegraphics[scale=0.2]{alpinist-climbing-peak-mountain-comic-hand-drawn-vector-illustration.jpg}
  \end{center}

@@

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their firstname as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I did my postdoc with Jeff Foster at Univ. of Maryland.   Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many international students include a lengthy title of people they know, e.g., I am advised by Asst. Prof. X, and I also collaborate with Distinguished Scientist Y.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. In short, these nuances represent some cultural differences that you may encounter and will gradually adapt to. More on cultural differences in \S\ref{sec:cultural}.
